---
title: "Exercise 3 --- Point Patterns in Space and Space-Time"
subtitle: "BSTA 374 · 25 points · Due Sunday, October 4, 2026, 11:59 pm"
author: "Abby Winters · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  warning: false
  message: false
---

## What this exercise is

The largest lab of the semester. You will decide whether a set of case locations is clustered, at what spatial
scale, where the intensity is highest, and whether the clustering is also
concentrated in *time* --- which is a different claim, and the one that matters
for an outbreak.

**Data.** Two files, posted with this exercise. Put them in a `data/` folder
beside your `.qmd`.

| File | What it holds |
|---|---|
| `dengue_cases_sim.csv` | 500 simulated dengue cases: `case_id`, `onset_date`, `report_date`, `neighborhood`, `age_group`, `lon`, `lat` |
| `study_area.geojson` | the study-area boundary polygon |

These are **simulated** data modelled on a dengue surveillance setting. Nobody's
health information is involved, and the generating process is known --- which
means your conclusions can be checked.

Each part ends with **starting code**: copy it, fill in the blanks marked
`______`, and adapt. It is scaffolding, not a solution.

::: {.callout-important}
## Submission
Two things. Submit `LASTNAME_FI_Exercise3.qmd` --- the source --- to Course Site,
with your name in the document header. And publish the rendered HTML to your
semester GitHub repository, then paste the live URL into the Course Site
submission --- Part H is that second step, and it is the one that carries the
point.

**Due Wednesday, October 1, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Restart R and render from a clean session before submitting. Several parts
involve simulation --- set a seed at the top and use it, or your numbers will
change between the render you checked and the render I grade.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(spatstat)
library(splancs)      # space-time K function
library(plotly)       # space-time cube

set.seed(800040125)      # your student ID works
```

---

## Part A --- From CSV to `ppp` (3 points)

Everything after this part runs on a `ppp` object, so it is worth knowing what
one is. `spatstat` does not work with `sf`. A **`ppp`** ("planar point pattern")
holds two things: the coordinates of the events, and a **window** --- the region
within which they could have been observed. Both are required, because every
statistic in this lab is a comparison against what would be expected inside that
window: the intensity is $n$ divided by its area, the expected nearest-neighbour
distance depends on that intensity, and the edge corrections depend on its shape.
Give `spatstat` the wrong window and every number below changes without any
warning.

1. Read the case file and the boundary. Report how many cases you have and what
   one row represents.
   
   **There are 500 dengue cases within the data set. Each row shows one reported dengue case, including its onset date, report date, neighborhood, age group and location coordinates.**
   
2. Convert the cases to an `sf` object in **EPSG:4326**, then project both the
   cases and the boundary to **EPSG:32618** (UTM 18N, metres). One sentence on
   why the analysis cannot be done in degrees.
   
   **Spatial analysis cannot be completed in degrees since distances, areas, and densities all require coordinate systems that are measured in linear units, like meters.**
   
3. Build a `ppp` whose window is the study-area polygon, then print it. Report
   the number of events and the intensity in cases per km².
   
   **There are 500 events with 2.63 cases per km^2 as the intensity. This means that on average about 2.6 dengue cases occur per square kilometer in the study area.**
   
4. Replace the polygon window with the bounding box of the cases
   (`Window(pp) <- owin(range(x), range(y))`) and recompute the intensity. Two or
   three sentences on how much it changed and why the choice of window is a
   modelling decision rather than a technicality. Then go back to the polygon.
   
   **Replacing the polygon window with the bounding box of cases, caused the intensity to go from 2.63 to 2.35 cases per km^2. The intensity decreased because the bounding box covers a larger area than the original polygon while both having the same 500 cases. This shows that the the estimated intensity depends on the study area that is chosen for the analysis and involves a modeling decision with the observation window. We can see that spatial intensity depends on the area at risk.**
   
5. Map the cases over the boundary.

### Starting code

```{r}
#| label: a-starter

cases_raw <- read.csv("dengue_cases_sim.csv", stringsAsFactors = FALSE)
area      <- st_read("study_area.geojson", quiet = TRUE) |> st_transform(32618)

cases <- cases_raw |>
  mutate(onset_date  = as.Date(onset_date),
         report_date = as.Date(report_date)) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32618)

win <- as.owin(st_geometry(area))       # the observation window
pp  <- as.ppp(st_geometry(cases))       # coordinates only, with a bounding box
Window(pp) <- win                       # attach the real window

pp                                      # print it: n, window type, extent
npoints(pp)
intensity(pp) * 1e6                     # cases per km^2

#A4
Window(pp) <- owin(range(pp$x), range(pp$y))
intensity(pp) * 1e6

#A5
plot(st_geometry(area))
plot(st_geometry(cases), add = TRUE, pch = 16, cex = 0.5)
```

---

## Part B --- Quadrat analysis (3 points)

1. Lay a 6 × 6 grid over the window, map the counts, and compute the mean,
   variance and **variance-to-mean ratio** from the counts by hand --- not from a
   canned function. Show the arithmetic.
   
   **VMR = 581.70159 / 13.88889 = 41.88251**
   
2. State what the VMR is being compared against, and why: under complete spatial
   randomness the counts are Poisson, and a Poisson distribution has variance equal to its mean. Report where your value sits relative to 1.
   
   **The VMR is being compared against 1 since under complete spatial randomness the counts are Poisson and with that, the variance equals the mean. A VMR of 1 would mean a random pattern and since the VMR is 41.88, it is much greater than 1 showing a strong cluster of dengue cases.**
   
3. Repeat the VMR for at least four grid resolutions and plot VMR against cells
   per side. Three or four sentences: what does the shape of that curve say about
   the claim "this pattern is clustered"?
   
   **The VMR values range from 26.65 to 41.41 and they are all well above 1 meaning that there is a strong cluster of dengue cases here rather than randomly distributed. Even though the values of the VMRs changed based on the grid, the conclusion still stayed the same (above 1) for all scales. Therefore, this reinforces that there is high evidence that the spatial distribution of cases is clustered.**

### Starting code

```{r}
#| label: b-starter

qc  <- quadratcount(pp, nx = 6, ny = 6)
cnt <- as.vector(qc)

c(k = length(cnt), mean = mean(cnt), var = var(cnt), VMR = var(cnt) / mean(cnt))

grid_n <- c(3, 4, 6, 8, 12)
vmr <- sapply(grid_n, function(k) {
  v <- as.vector(quadratcount(pp, nx = k, ny = k))
  var(v) / mean(v)
})

#B3 
data.frame(grid_n, vmr)
```

---

## Part C --- Nearest-neighbour statistics (3 points)

1. Compute `nndist()`, report the observed mean, and compute the CSR expectation
   $E(d) = 1/(2\sqrt{\lambda})$ **by hand** from your intensity. Report the
   Clark--Evans ratio $R$ and state the direction of the departure.
   
   **The observed mean is 215.25 m and the CSR expectation is 308.50 m. The Clark-Evans Ratio is about R = .6985 and this is less than 1 meaning that the points are closer together than expected under CSR, showing there is a clustered spatial pattern.**
   
2. Run `clarkevans.test()` with a Monte Carlo null. The window is an irregular
   polygon, so the Donnelly correction is unavailable --- use
   `correction = "cdf"` with `method = "MonteCarlo"`. One sentence on why an edge
   correction is needed at all.
   
   **The Clark-Evans test returned R = .65 with p-value = .01. The p-value is less than .05 so we reject the null hypothesis meaning that there is enough evidence that the cases are not randomly distributed. This means that they are most likely clustered. An edge correction is crucial because points near the boundary of the study area have less neighboring locations around them and this can cause bias to the nearest-neighbor distances.** 
   
   
3. Plot the G function with a simulation envelope. Two or three sentences on what
   G shows that the single ratio $R$ does not.
   
   **The G function shows how clustering changes across a range of different distances and the Clark-Evans ratio shows the pattern with a single value. In the plot, the G function lies is above the simulation envelope at shorter distances, meaning that cases have more nearby neighbors than would be expected under complete spatial randomness. This means that dengue cases are clustered at short distances.** 

### Starting code

```{r}
#| label: c-starter

nnd    <- nndist(pp)
lambda <- intensity(pp)
d_exp  <- 1 / (2 * sqrt(lambda))  # 1 / (2 sqrt(lambda))

c(observed = mean(nnd), expected = d_exp, R = mean(nnd) / d_exp)

clarkevans.test(pp, correction = "cdf", alternative = "clustered",
                method = "MonteCarlo", nsim = 99)

plot(envelope(pp, Gest, nsim = 39, verbose = FALSE), main = "", legend = FALSE)
```

---

## Part D --- Ripley's K across scales (4 points)

1. Estimate $K(r)$ with an edge correction, then plot the centered L function
   $\hat{L}(r) - r$ against $r$.
2. Add a **simultaneous** envelope (`global = TRUE`) from at least 99 CSR
   simulations. State the distances over which the curve leaves the band.
   
   **The L(r)-r curve lies above the envelope for almost all the distances shown. This means spatial clustering over all scales in the study area.** 
   
3. Read a **characteristic scale** off the curve --- the $r$ at which the
   departure peaks --- and translate it into something a public health officer
   would recognise: a few city blocks, a neighbourhood, half the city.
   
   **The peak is around r = 800 to 900. The means that clustering is strongest at a distance of 800-900 meters, referring to the scale of a neighborhood.** 
   
4. Three or four sentences: $K$ assumes the intensity is constant across the
   window. Looking at your map from Part A, is that plausible here? Say what the
   consequence is for reading the curve --- what else, besides interaction
   between cases, could put it above the envelope.
   
   **K assumes a constant intensity across the study area. This might not be realistic since some neighborhoods could have more cases than others. The clustering shown by the curve could be a result of population density or other trends, not just only interaction between cases.**

::: {.callout-warning}
`global = TRUE` with `nsim = 99` takes a minute or two. Develop with
`nsim = 19`, raise it once, and do not re-render in a loop.
:::

### Starting code

```{r}
#| label: d-starter

L <- Lest(pp, correction = "isotropic")
plot(L, . - r ~ r, main = "", legend = FALSE)

env <- envelope(pp, Lest, nsim = 99, correction = "isotropic",
                global = TRUE, verbose = FALSE)
plot(env, . - r ~ r, main = "", legend = FALSE)

```

---

## Part E --- Kernel density estimation and bandwidth choice (4 points)

K tells you *whether* and *at what distance*; it does not tell you *where*. That
is what a kernel intensity surface is for.

1. Estimate the intensity with `density.ppp()` and convert to **cases per km²**.
   Map it with the cases drawn on top.
2. Produce the same surface at **three bandwidths** --- one clearly too small,
   one clearly too large, and one you would defend --- on a **common colour
   scale** so the panels are comparable. Two or three sentences on what changes
   and what does not.
   
   **The smallest bandwidth, 84.35 m, produces a detailed and patchy surface with a few peaks. The selected bandwidth, 168.7 m, shows the data much smoother while still keeping the same peaks. The third bandwidth, 337.39 m, makes the data extremely smooth and reduces the detail. The peaks are still visible though.** 
   
3. Compare at least two automatic selectors (`bw.diggle()`, `bw.ppl()`,
   `bw.scott()`). Report the values in metres, say which you adopt and why, and
   note whether your answer to step 2 would change if you adopted a different one.
   
   **The automatic bandwidth selectors returned values of 168.7 m (Diggle), 469.93 m (PPL), and 1250.55 m (Scott). I adopted the Diggle bandwidth because it preserves the patterns, shows the peaks clearly, and doesn't make the surface too rough or smooth. Using PPL or Scott would produce a smoother surface, but the peaks would remain in the same spot.** 
   
4. Three or four sentences on what this map is and is not. Where is the intensity
   highest? What would you need in order to turn this surface into a statement
   about **risk** rather than about counts, and what is the name of the quantity
   you would compute instead?
   
   **This map shows the estimated spatial intensity of dengue cases in cases per km squared. It shows where cases are concentrated, but it does not show or measure the risk of infection. The highest intensity occurs in several different spots of the study area, specifically east, west, and north. To make statements about risk, population data would need to be collected so that cases per population could be calculated instead of case intensity only. **

::: {.callout-tip}
## What earns full marks here
Step 4, and the common colour scale in step 2. Two surfaces at different
bandwidths with independently stretched legends look like different data; a
reader cannot compare them, and neither can you.
:::

### Starting code

```{r}
#| label: e-starter

sig <- bw.diggle(pp)
c(diggle = bw.diggle(pp), ppl = bw.ppl(pp), scott = mean(bw.scott(pp)))

kde <- density(pp, sigma = sig, eps = 50) * 1e6       # cases per km^2
plot(kde, main = "", ribargs = list(las = 1))
plot(Window(pp), add = TRUE, border = "grey40")
points(pp, pch = 16, cex = 0.25, col = rgb(0, 0, 0, 0.35))

sigmas <- c(sig/2, sig, 2*sig)                   # too small, defensible, too large
surf   <- lapply(sigmas, function(s) density(pp, sigma = s, eps = 50) * 1e6)
zmax   <- max(sapply(surf, max))                      # one scale for all panels

par(mfrow = c(1, 3))
for (i in seq_along(surf))
  plot(surf[[i]], main = paste0("sigma = ", sigmas[i], " m"), zlim = c(0, zmax))
```

---

## Part F --- Space-time kernel density (3 points)

A single intensity surface pools six months. An outbreak is a hotspot that
exists in some weeks and not others, and pooling hides exactly that.

1. Split the cases into **monthly** patterns and estimate a kernel surface for
   each, using **one bandwidth for all months** and **one colour scale for all
   months**. Display them as small multiples.
2. Two or three sentences on why both of those constraints are necessary, and
   what a reader would wrongly conclude if each panel were scaled to its own
   maximum.
   
   **Using one bandwidth for all months makes sure that differences between maps shows changes in the data rather than changes in smoothing. Using one color scale for all months lets intensity values to be compared directly. If each month used its own different scale, weak hotspots could come up as strong and it could be interpreted incorrectly.** 
   
3. Name the months and locations where the intensity is high, and say which
   hotspots persist across months and which appear and disappear.
   
   **Highest intensity occurs in February in the north area. March and May are also hotspots. These peals change location from month to month instead of staying in the same place. No hotspot appears to remain throughout the whole study period. This means that dengue cases and activity shift spatially over time.** 
   
4. Build a **space-time cube**: an interactive 3D scatter with easting, northing
   and day-of-study on the three axes, coloured by neighbourhood. Rotate it until
   the structure is visible, then set a viewing angle that makes your point and
   keep it. One or two sentences on what the cube shows that the small multiples
   do not, and vice versa --- look in particular for the difference between a
   compact blob and a vertical column.
   
   **The space-time cube show location and timing of dengue cases. Compact clusters show outbreaks that occurred in a certain time and place, while vertical columns show cases occurring in the same area over a longer period. The monthly map shows where hotspots occur, but the cube shows when they occur as well.** 

::: {.callout-warning}
## plotly renders in HTML only
The cube is an HTML widget: it appears in an HTML render and is silently dropped
from a PDF one. The two chunks below are gated on `knitr::is_html_output()`, so
whichever format you render gets a working figure. If you publish the HTML, keep
`embed-resources: true` --- otherwise the widget lives in a `_files/` folder that
has to travel with the page.
:::

::: {.callout-note}
## If you want the proper estimator
Steps 1--3 are a discretised space-time KDE: one spatial bandwidth, and a
temporal "bandwidth" equal to one month, applied as a hard cut. The continuous
version smooths in time as well, with a separate temporal bandwidth, and lives
in the `sparr` package --- `spattemp.density()` with `h` and `lambda`, then
`spattemp.slice()` to pull out time slices. Using it is worth up to **2 bonus
points** if the monthly version is already complete and correct. Do not start
here.
:::

### Starting code

```{r}
#| label: f-starter

cases$month <- format(cases$onset_date, "%Y-%m")
h <- sig                                  # one bandwidth, chosen in Part E

by_month <- lapply(sort(unique(cases$month)), function(m) {
  sub <- cases[cases$month == m, ]
  p   <- ppp(st_coordinates(sub)[, 1], st_coordinates(sub)[, 2], window = Window(pp))
  density(p, sigma = h, eps = 50) * 1e6
})

zmax <- max(sapply(by_month, max))        # common scale across months

par(mfrow = c(2, 3), mar = c(1, 1, 2, 2))
for (i in seq_along(by_month))
  plot(by_month[[i]], main = sort(unique(cases$month))[i], zlim = c(0, zmax))

xy  <- st_coordinates(cases)
cube <- data.frame(E = xy[, 1], N = xy[, 2],
                   day = as.numeric(cases$onset_date - min(cases$onset_date)),
                   hood = cases$neighborhood)

#| label: f-cube-html
#| eval: !expr knitr::is_html_output()

library(plotly)

plot_ly(cube, x = ~E, y = ~N, z = ~day, color = ~hood,
        type = "scatter3d", mode = "markers",
        marker = list(size = 3, opacity = 0.75)) |>
  plotly::layout(scene = list(xaxis = list(title = "easting (m)"),
                      yaxis = list(title = "northing (m)"),
                      zaxis = list(title = "day of study")))
```

```{r}
#| label: f-cube-static
#| eval: !expr isFALSE(knitr::is_html_output())

library(scatterplot3d) #Needed to install and load package to use below 
scatterplot3d::scatterplot3d(cube$E, cube$N, cube$day, pch = 16, cex.symbols = 0.4,
                             color = as.integer(factor(cube$hood)),
                             xlab = "easting (m)", ylab = "northing (m)", zlab = "day")
```

---

## Part G --- The space-time K function (4 points)

Spatial clustering and space-time clustering are different claims. A
neighbourhood with chronically high incidence is clustered in space in every
month; an outbreak is clustered in space *and* time together. Diggle's space-time
K function separates them by comparing the joint function with the product of the
two marginals:

$$
D(s, t) = K_{st}(s, t) - K_s(s)\,K_t(t)
$$

$D = 0$ means space and time are independent --- cases may cluster spatially,
temporally, or both, but not *jointly*. $D > 0$ is space-time interaction.

1. Build the inputs `splancs` expects: a two-column matrix of projected
   coordinates, a numeric time vector (days since the first onset), and the
   boundary as a matrix of vertices.
2. Compute $K_s$, $K_t$ and $K_{st}$ with `stkhat()`. Justify your choice of
   spatial distances `s` and time lags `tm` in one sentence each --- what scales
   are plausible for a mosquito-borne disease?
   
   **Spatial distances from 50 m- 3000 m were used because dengue transmission is likely to occur over neighborhood scaled distances within a city. Time lags from 3-70 days were used since they cover the period during which cases may be linked through mosquito transmission and the incubation period of the disease. ** 
   
3. Plot **both** $D(s,t)$ and the standardised version
   $D_0(s,t) = K_{st}/(K_s K_t) - 1$. One or two sentences on why the two
   surfaces have different shapes: $K$ functions are cumulative, so $D$
   accumulates and then flattens, while $D_0$ measures the *proportional* excess
   and falls away once you leave the scale of the clustering. This is the same
   move as plotting $\hat{L}(r) - r$ instead of $\hat{K}(r)$ in Part D.
   
   **The D(s,t) surface increases then levels off, while the D0(s,t) surface is highest at small distances and time lags and then slowly decreases. The standardized surface makes the strongest space-time clustering easier to identify because it emphasizes relative excess clustering rather than the cumulative total.** 
   
4. A surface is hard to read against nothing. Build a **permutation envelope**:
   shuffle the onset dates across the fixed locations, recompute $D_0$, repeat 99
   times, and take the upper 97.5% quantile at each $(s, t)$. Then plot $D_0$
   against $s$ as a curve per time lag, with its envelope --- the space-time
   analogue of the envelope you used for $L$. Shuffling the dates destroys any
   space-time interaction while leaving both marginal patterns exactly as
   observed, which is the correct null here.
5. Read the peak off those curves --- a distance and a lag --- and say what the
   pair means for this disease.
   
   **The strongest peak occurs at short distances (about 0-500 m) and short time lags (about 3-9 days). This shows that dengue cases that occur close also occur close in time, which makes sense with localized disease transmission.**
   
6. Run `stmctest()` for a single global p-value. It sums the residuals
   $\hat{D}(s,t)$ over the whole grid into one statistic $U$, recomputes it under
   permuted dates, and returns the observed value as `$t0` and the permuted ones
   as `$t`. Report the p-value as
   $(1 + \#\{t \geq t_0\}) / (1 + B)$ and plot the null distribution with the
   observed value marked. One sentence on what this test adds to the curves in
   step 4, and what it cannot tell you.
   
   **The p-value is .001. This shows strong evidence against the null hypothesis of no space-time interaction. The test summarizes evidence across space-time surface into a single number, but it not show the specific distances or time lags where cluestering is strongest. ** 
   
7. Five or six sentences. Tabulate cases by neighbourhood and month. At least one
   neighbourhood contributes to the spatial clustering of Part D without
   contributing to the space-time interaction here. Identify it, explain what
   distinguishes it from the others, and say what each of the two patterns would
   mean for someone deciding where to send a vector-control team this week.
   
   **El Retiro contributes to the spatial clustering observed in Part D because it consistently shows a relatively large number in dengue cases. However, the cases are spread across all 6 months rather than being concentrated in a short time period, so it contributes less to the space-time interaction seen in Part G. On the other hand, neighborhoods such as Aguablanca, Floralia, and Siloe experienced large spikes of cases within a single month, meaning localized outbreaks in space and time. Spatial clustering shows places where many cases occur overall, while space-time interaction shows places where cases occur in close proximity of both space and time. For a vector-control team planning where to work this week, spatial clustering would highlight neighborhoods with persisitently high case counts, whereas space-time interaction would help show active outbreaks that require immediate attention.  **

::: {.callout-warning}
## Never report $p = 0$
`mean(mc$t >= mc$t0)` returns 0 when no permutation beats the observed value,
and 0 is not a p-value --- with $B$ permutations you cannot resolve anything
below $1/(B+1)$. Always add the observed value to both counts, as above. With
999 permutations the smallest reportable value is 0.001, and the honest phrasing
is "$p \leq 0.001$", not "$p < 0.0001$".
:::

::: {.callout-note}
`splancs` is old and its arguments are terse: `stkhat(pts, times, poly, tlimits,
s, tm)`, where `tlimits` is the time range, `s` the spatial distances and `tm`
the time lags. `?stkhat` and `?stmctest` are short. A **Knox test** on a distance
and time threshold of your choosing is an acceptable substitute if you defend
both thresholds --- say explicitly if you go that route.
:::

### Starting code

```{r}
#| label: g-starter

xy    <- st_coordinates(cases)
pts   <- as.points(xy[, 1], xy[, 2])
times <- as.numeric(cases$onset_date - min(cases$onset_date))
bdry  <- st_coordinates(area)[, 1:2]         # drop the closing vertex if duplicated

s  <- c(seq(50, 1000, by = 50), seq(1200, 3000, by = 200))   # finer near the origin
tm <- seq(3, 70, by = 3)                     # days

stk <- stkhat(pts, times, bdry, c(0, max(times)), s, tm)

D  <- stk$kst - outer(stk$ks, stk$kt)                  # raw excess
D0 <- stk$kst / outer(stk$ks, stk$kt) - 1              # standardised excess

par(mfrow = c(1, 2))
persp(s, tm, D,  theta = 30, phi = 25,
      xlab = "distance (m)", ylab = "lag (days)", zlab = "D")
persp(s, tm, D0, theta = 30, phi = 25,
      xlab = "distance (m)", ylab = "lag (days)", zlab = "D0")

# permutation envelope: shuffle the dates, keep the locations
B <- 99
perm <- replicate(B, {
  st <- stkhat(pts, sample(times), bdry, c(0, max(times)), s, tm)
  st$kst / outer(st$ks, st$kt) - 1
})
hi <- apply(perm, c(1, 2), quantile, 0.975)

k <- c(1, 3, 7, 14)                          # lags to show
matplot(s, D0[, k], type = "l", lty = 1, lwd = 2, xlab = "distance s (m)",
        ylab = expression(D[0](s, t)))
matlines(s, hi[, k], lty = 3, lwd = 1)
abline(h = 0, col = "grey")
legend("topright", legend = paste("lag <=", tm[k], "d"), lty = 1, lwd = 2,
       col = 1:4, bty = "n")

mc <- stmctest(pts, times, bdry, c(0, max(times)), s, tm, nsim = 999, quiet = TRUE)

(1 + sum(mc$t >= mc$t0)) / (1 + length(mc$t))   # global Monte Carlo p-value

hist(mc$t, xlim = range(c(mc$t, mc$t0)), xlab = "U under permuted dates", main = "")
abline(v = mc$t0, col = "firebrick", lwd = 2)

table(cases$neighborhood, cases$month)
```

---

## Part H --- Rendering and publication (1 point)

The `.qmd` is the submission, but the rendered page is what anyone else can
actually read --- and by the end of the semester this repository is your
portfolio.

1. Render to **HTML** from a clean session, with the seed set and every package
   loaded explicitly, then commit both the `.qmd` and the rendered output to your
   semester repository, in the folder your site is built from.
2. Push, wait for GitHub Pages to rebuild, and **open the published URL yourself**
   to check it. A page that renders locally and 404s online is not submitted.
3. Put the URL of the published page in your document header and in your Course
   Site submission.

The point is for a page that works: rendered, pushed, and reachable at the URL
you supply.

::: {.callout-warning}
## Two things that break the published page
`embed-resources: true` is already set in the header --- keep it. Without it the
`plotly` cube in Part F lives in a `_files/` folder that must be committed
alongside the HTML, and forgetting it gives a page with a blank space where the
cube should be.

Relative paths to `data/` work locally because your working directory is the
`.qmd`'s folder. They will also work on the site *as long as you commit the data
files*. Check the live page, not the local preview.
:::

---

## AI use {.unnumbered}

End your document with a short **AI use note**: which tools, what for, at which
stage. Two or three sentences.

AI use is unrestricted in this course and costs you nothing, but you must be able
to explain every line you hand in. The note carries no points --- it is simply
expected, the way a methods section is.


**I used AI when library(spatstat) wouldn't load. It advised I needed to install it within the terminal and then it worked. I got an error between lines 412-422 when trying to render and I couldn't figure out what was the issue so I used AI to help debug it. First, it was giving me an error with the plotly package so I readded the package. Then, it gave me an error saying cube was not defined even though it was in the code chunk above. With AI's help, it suggested I put the two chunks into one. When I did that, it allowed me to render. ** 

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct import, CRS and window; `ppp` built and explained; effect of the window argued | 3 |
| **B** | VMR computed from counts; benchmark of 1 explained; resolution sweep interpreted | 3 |
| **C** | $E(d)$ derived and $R$ interpreted; Monte Carlo test correct; G compared with $R$ | 3 |
| **D** | Centred L with a simultaneous envelope; characteristic scale read and translated; stationarity assumption discussed | 4 |
| **E** | Surface in interpretable units; three bandwidths on a common scale; selectors compared; density-is-not-risk argument | 4 |
| **F** | Monthly surfaces with common bandwidth and scale; persistent vs transient hotspots named; space-time cube | 3 |
| **G** | Space-time K computed; $D_0$ with a permutation envelope; peak distance and lag read off; the spatial-only neighbourhood identified and the operational difference explained | 4 |
| **H** | Clean render to HTML, published to GitHub Pages; working URL supplied | 1 |
| | **Total** | **25** |

Writing counts for roughly 15% within each part --- more than in Exercise 1,
because more of this lab is interpretation. A submission that is all code and no
sentences cannot score above about 50%.

---

## Getting help

- Office hours, or email with the code and the exact error message.
- `splancs` in Part G is the one place where the documentation is thin. Start
  early enough that you can ask.
- Stuck for more than thirty minutes on something mechanical: ask rather than
  persevere.
